---
title: "Text as Data Workshop"
author: "Siefken"
date: "2026-05-25"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
subtitle: SCISS Sharjah 2026
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

# Setup

## Load Packages

```{r install-packages, eval=FALSE}
install.packages("pacman")
```

```{r load-packages}
library(pacman)

p_load(tidyverse,
       quanteda,
       quanteda.textmodels,
       quanteda.textplots,
       quanteda.textstats,
       stm,
       devtools,
       SnowballC)
```

## Install & Load Additional Packages from GitHub

Download Presidential Debate Dataset from Github (Martherus 2020; <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3611815>)

```{r install-github-packages, eval=FALSE}
devtools::install_github("jamesmartherus/debates")
devtools::install_github("kbenoit/quanteda.dictionaries")
devtools::install_github("quanteda/quanteda.sentiment")
```

```{r load-extra-packages}
library("quanteda.sentiment")
library("quanteda.dictionaries")
library(debates)
```

## Load the Data

```{r load-data}
data(debate_transcripts)
```

---

# Data Pre-Processing

```{r preprocessing}
debates <- debate_transcripts %>% 
  filter(candidate == 1) %>% #Only include candidates
  filter(election_year >= 2000, type == "Pres") %>% #Limit the data
  group_by(speaker, date) %>%
  summarise(text = paste(text, collapse = " "), #Merge all speeches per candidate and day into one text
            election_year = first(election_year), #Needed to keep the meta data
            type = first(type),
            .groups = "drop") %>%
  mutate(doc_id = row_number())
```

## Create Corpus

```{r create-corpus}
corpus_debate <- corpus(debates,
                      text_field = "text",
                      docid_field = "doc_id")
```

## Overview of the Data

```{r corpus-summary}
summary_debate <- summary(corpus_debate, n = ndoc(corpus_debate)) #By default, summary only shows the first 100 rows
```

```{r view-corpus-summary, eval=FALSE}
View(summary_debate)
```

```{r plot-tokens}
summary_debate %>%
  ggplot(aes(x = speaker, y = Tokens)) +
  geom_boxplot(aes(fill = speaker), alpha = .7) +
  theme_bw()
```

## Subsetting the Corpus

```{r subset-corpus}
corpus_obama <- corpus_subset(corpus_debate, 
                              speaker == "Barack Obama")
```

---

# Document Feature Matrix

Turn dataset into a DFM, required for e.g. Dictionary Analysis.

```{r create-dfm}
dfm_debate <- corpus_debate %>%
  tokens(
    remove_punct = TRUE,
    remove_numbers = TRUE,
    split_hyphens = TRUE
  ) %>%
  tokens_wordstem(
    language = quanteda_options("language_stemmer"),
    verbose = quanteda_options("verbose")
  ) %>%
  tokens_remove(c("example1", "example2", stopwords("en"))) %>% #Remove specified words
  dfm()
```

## Trim Infrequent Features

```{r trim-dfm}
dfm_debate_trim <- dfm_trim(dfm_debate, 
                          min_docfreq = 2, #Has to appear in at least two texts
                          min_termfreq = 5) #Has to appear at least 5 times total

nfeat(dfm_debate_trim) #Number of unique words in the data
```

## Inspect the DFM

```{r view-dfm, eval=FALSE}
View(convert(dfm_debate, to = "data.frame"))
```

---

# Analyse the DFM

## Top 20 Features (Overall)

```{r top-features}
top20_feat <- textstat_frequency(dfm_debate_trim, n = 20)
```

```{r view-top-features, eval=FALSE}
View(top20_feat)
```

## Top 20 Features by Speaker

```{r top-features-by-speaker}
docvars(dfm_debate_trim)

top20_feat_president <- textstat_frequency(dfm_debate_trim, 
                                 n = 20,
                                 group = docvars(dfm_debate_trim, "speaker"))
```

```{r view-top-features-by-speaker, eval=FALSE}
View(top20_feat_president)
```

---

# Wordclouds

```{r wordcloud-basic}
textplot_wordcloud(dfm_debate_trim)
```

```{r wordcloud-custom}
textplot_wordcloud(dfm_debate_trim,
                   max_words = 50,
                   color = "red",
                   rotation = 0)
```

---

# Keyness Analysis

Keyness analysis determines which words are used more by a predetermined group than by the other groups. E.g. which words are used more by Trump than other candidates.

```{r keyness}
#Lets group the DFM by Candidates
dfm_debate_grouped <- dfm_debate_trim %>%
  dfm_group(groups = docvars(dfm_debate_trim, "speaker"))

keyness_trump <- textstat_keyness(dfm_debate_grouped, target = "Donald Trump")

textplot_keyness(keyness_trump,
                 show_reference = FALSE,
                 n = 25)
```

---

# Dictionary Analysis

## Create a Custom Dictionary

Note: As we used stemming on our data, the dictionary should also contain stemmed words.

```{r stemming-example}
wordStem("integration", language = "en")
```

```{r create-dictionary}
list_dict <- list(
  economy = c("economi", "job", "employ", "gdp", "growth", "budget", "deficit", "debt"),
  healthcare = c("health", "medicaid", "medicar", "insur", "hospit", "doctor", "patient"),
  education = c("educ", "school", "teacher", "student", "colleg", "univers", "tuition"),
  environment = c("climat", "environ", "energi", " green", "emiss", "renew", "pollut"),
  immigration = c("immigr", "border", "citizen", "visa", "migrant", "deport"),
  foreign_policy = c("foreign", "diplomat", "diplomaci", "allianc", "nato", "trade", "tariff", "sanction"),
  social_security = c("retir", "pension", "senior", "elder", "social secur"),
  crime = c("crime", "crimin", "polic", "justic", "prison", "law")
)

topic_dict <- dictionary(list_dict)
```

## Apply the Dictionary

```{r apply-dictionary}
topic_counts <- dfm_lookup(dfm_debate_trim,
                           topic_dict)

df_topic_counts <- convert(topic_counts,
                           to = "data.frame")

df_topic_counts <- cbind(docvars(dfm_debate_trim), df_topic_counts) #Merge meta info back on the dataset
```

## Summarise by Speaker

```{r summarise-topics}
df_topic_counts %>%
  group_by(speaker) %>%
  summarise(mean_immigration = mean(immigration),
            mean_economy = mean(economy),
            mean_environment = mean(environment),
            mean_crime = mean(crime))
```

---

# Sentiment Analysis

Sentiment Dictionary from the `quanteda.sentiment` package (Benoit, <https://github.com/quanteda/quanteda.sentiment>).

For this example, we will use the Lexicoder Sentiment Dictionary by Young & Soroka (2012).

```{r view-sentiment-dict, eval=FALSE}
View(data_dictionary_LSD2015)
View(list_dict)
```

```{r sentiment-analysis}
debate_sent <- dfm_lookup(dfm_debate_trim,
                        data_dictionary_LSD2015)

debate_sent_df <- convert(debate_sent,
                           to = "data.frame")

debate_sent_df <- cbind(docvars(dfm_debate_trim),
                      debate_sent_df)

debate_sent_df <- debate_sent_df %>%
  mutate(sentiment_score = (positive - negative)/(positive+negative),
         perc_pos = positive/(positive+negative),
         perc_neg = negative/(positive+negative))
```

```{r view-sentiment-df, eval=FALSE}
View(debate_sent_df)
```

## Visualise Sentiment Over Time

```{r plot-sentiment-score}
ggplot(debate_sent_df, aes(x = date, y = sentiment_score)) +
  geom_point() +
  geom_smooth()
```

```{r plot-positive}
ggplot(debate_sent_df, aes(x = date, y = perc_pos)) +
  geom_point() +
  geom_smooth()
```

```{r plot-negative}
ggplot(debate_sent_df, aes(x = date, y = perc_neg)) +
  geom_point() +
  geom_smooth()
```

---

# Topic Models

For this, we will use the STM package (Roberts 2025, <https://cran.r-project.org/web/packages/stm/index.html>).

```{r stm-convert}
dfm_stm <- convert(dfm_debate_trim, to = "stm", docvars(dfm_debate_trim))
```

## Fit the STM Model

Now we specify how many topics we expect within the data by defining `K`. We tell STM to run only 50 iterations max — skip this argument in a real analysis.

```{r stm-model,results='hide'}
stm_model <- stm(documents = dfm_stm$documents,
                 vocab = dfm_stm$vocab,
                 K = 6,
                 prevalence = ~ stm::s(election_year), #We allow the prevalence of topics to vary by year and speaker
                 max.em.its = 50,
                 data = dfm_stm$meta,
                 seed = 26052026)
```

## Inspect Topics

Common rule: Look at the FREX words as the most frequent and exclusive words per topic.

```{r label-topics}
labelTopics(stm_model)

topic_labels <- c("2016 Debate",
                  "Social Policy",
                  "Other Candidates",
                  "Iraq War",
                  "Tax & Labor",
                  "Foreign Policy")
```

## Plot Topic Model Results

```{r stm-plot-summary}
plot(stm_model, type = "summary",
     labeltype = "frex",
     custom.labels = topic_labels)
```

```{r stm-plot-hist}
plot(stm_model, type = "hist",
     labeltype = "frex",
     custom.labels = topic_labels)
```

## Estimate Effects

```{r estimate-effects}
prep <- estimateEffect(~ stm::s(election_year),
                       stm_model,
                       meta = dfm_stm$meta)

par(mfrow = c(1,1))
plot(prep, covariate = "election_year", 
     topics = 1:6,
     model = stm_model,
     method = "continuous",
     xlab = "Election Year",
     custom.labels = topic_labels,
     labeltype = "custom")
```

---

# Wordfish

```{r wordfish}
wfish <- textmodel_wordfish(dfm_debate_trim)
```

## Plot Document Positions

```{r wordfish-documents}
textplot_scale1d(wfish, margin = "documents")
```

## Plot Feature Positions

```{r wordfish-features}
textplot_scale1d(wfish, margin = "features")
```

## Save Theta Scores

```{r wordfish-save}
docvars(dfm_debate_trim, "theta") <- wfish$theta
```

---

# Wordscores

We pre-define one text by each Democrat candidate as `1` and for Republicans as `-1`.

```{r wordscores}
dfm_debate_trim$refvalue <- c(1,rep(NA,2), #Al Gore
                              1,rep(NA,5), #Obama
                              -1,rep(NA,4),#Trump
                              -1,rep(NA,5),#Bush
                              1,rep(NA,2), #Hillary
                              1,NA,        #Biden
                              1,rep(NA,2), #Kerry
                              -1,rep(NA,2),#McCain
                              -1,rep(NA,2))#Romney

wscore <- textmodel_wordscores(dfm_debate_trim, dfm_debate_trim$refvalue)

lbg <- predict(wscore, se.fit = TRUE,
               interval = "confidence",
               rescaling = "lbg")
```

## Visualise Scores per Text

```{r wordscores-plot}
textplot_scale1d(lbg,
                margin = "documents",
                groups = dfm_debate_trim$speaker)
```

## Save and Plot Wordscores Positions

```{r wordscores-save-plot}
positions_ws <- lbg$fit %>% as.data.frame()

docvars(dfm_debate_trim) <- bind_cols(docvars(dfm_debate_trim),
                                    positions_ws)

ggplot(docvars(dfm_debate_trim), 
       aes(x = fit,
           y = reorder(paste(speaker, date), fit))) +
  geom_point()
```

---

*Thank you for your attention!*
